Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
stuff

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
overloading extensively.

@@

In this section, we introduce the syntax of type theory with function
space and universes in a minimalistic way, only introducing operations
that are unavoidable. We eschew boilerplate by only defining
well-formed, well-scoped, well-typed terms that are quotiented by
conversion. This means that the equalities that hold between terms are
the ways we can convert terms to each other when running them as
programs. We give a tutorial-style introduction to the syntax, we do
not assume prior knowledge of the metatheory of type theory. The
definition of type theory that we obtain is the same as the single
substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
SOGAT (\ref{eq:tt}).

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists and not
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{alignat*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\begin{alignat*}{10}
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\end{alignat*}
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index, we denote it by $\q :
\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
such an $A'$. For this, we will introduce an operation $\blank[\p] :
\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
of introducing just $\blank[\p]$, we generalise a bit and add a new
sort $\Sub$ which for now only has the single element $\p$ and the
operation $\blank[\blank]$ is called \emph{instantiation}. We
instantiate the type with the weakening in
$\Sub\,\Delta\,\Gamma$. Elements of $\Sub$ will be in general called
substitutions, hence the name, but for now, we only have single end-of
context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
the codomain of a $\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
& \Sub && : \Con\ra\Con\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices \cite{debruijn}): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
new kind of weakening which adds a new variable in the last but one
position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\begin{alignat*}{10}
  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\end{alignat*}
where $+^n$ means the $n$-times iteration of $\blank^+$.

@@

So far our only terms are variables, but we would like to define
functions via lambda abstraction. Abstraction takes a term in a
context extended by the domain of the function. It comes with a rule
for instantiation analogous to $\Pi[]$.
\begin{alignat*}{10}
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
\end{alignat*}
Note that the equation $\lam[]$ only makes sense because of the
previous equation $\Pi[]$: the left hand side is in
$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
right hand side is in
$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.

The functions $\lambda x.x$ and $\lambda x\,y.x$ can be defined in our
syntax as $\lam\,\q : \Tm\,\Gamma\,(A\Ra A)$ and
$\lam\,(\lam\,(\q[\p])) :
\Tm\,\Gamma\,\big(\Pi\,A\,(B\Ra(A[\p]))\big)$ which make sense for all
$\Gamma$, $A$, $B$.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the typal version of the
equation mentioned in the introduction.
\begin{alignat*}{10}
& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the first-order single substitution semantics of the
SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each rule has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules. For example, we show how to
extend Definition \ref{def:tt} with the SOGAT-rules (\ref{eq:sigma}).
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules for $\Sigma$ into one isomorphism, and did not list
  substitution laws for $\fst$ and $\snd$ as they are derivable.
\end{definition}

>>>>

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way. We present
  our calculus in an easy-to-understand way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
overloading extensively, for example $\blank[\blank]$ for types and
terms overloaded.

@@

In this section, we introduce the syntax of type theory with function
space and universes in a minimalistic way, only introducing operations
that are unavoidable. We eschew boilerplate by only defining
well-formed, well-scoped, well-typed terms that are quotiented by
conversion. This means that the equalities that hold between terms are
the ways we can convert terms into each other when running them as
programs. We give a tutorial-style introduction to the syntax, we do
not assume prior knowledge of the metatheory of type theory. The
definition of type theory that we obtain is the same as the single
substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
SOGAT (\ref{eq:tt}).

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{alignat*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\begin{alignat*}{10}
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\end{alignat*}
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index \cite{debruijn}, we denote it by $\q :
\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
such an $A'$. For this, we will introduce an operation $\blank[\p] :
\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
of introducing just $\blank[\p]$, we generalise a bit and add a new
sort $\Sub$ which for now only has the single element $\p$ and the
operation $\blank[\blank]$ is called \emph{instantiation}. We
instantiate the type with the $\Sub\,\Delta\,\Gamma$ weakening.
Elements of $\Sub$ will be later called
substitutions, hence the name, but for now, we only have single end-of
context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
the codomain of a $\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
& \Sub && : \Con\ra\Con\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
new kind of weakening which adds a new variable in the last but one
position. But after introducing $\p'$, we would need another equation computing $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\begin{alignat*}{10}
  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\end{alignat*}
where $+^n$ means the $n$-times iteration of $\blank^+$.

@@

So far our only terms are variables, but we would like to define
functions via lambda abstraction. Abstraction takes a term in a
context extended by the domain of the function. It comes with a rule
for instantiation analogous to $\Pi[]$.
\begin{alignat*}{10}
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
\end{alignat*}
Note that the equation $\lam[]$ only makes sense because of the
previous equation $\Pi[]$: the left hand side is in
$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
$b[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
right hand side is in
$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.

The functions $\lambda x.x$ and $\lambda x\,y.x$ can be defined in our
syntax as $\lam\,\q : \Tm\,\Gamma\,(A\Ra A)$ and
$\lam\,(\lam\,(\q[\p])) :
\Tm\,\Gamma\,\big(\Pi\,A$ $(B\Ra(A[\p]))\big)$ which make sense for all
$\Gamma$, $A$, $B$.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the typal version of the
equation mentioned in the introduction (we don't have its term version).
\begin{alignat*}{10}
& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the first-order single substitution semantics of the
SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules. For example, we show how to
extend Definition \ref{def:tt} with the SOGAT-rules (\ref{eq:sigma}).
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules for $\Sigma$ into one isomorphism, and did not list
  substitution laws for $\fst$ and $\snd$ as they are derivable.
\end{definition}
